Concurrent Programming with CXL

### 2020-06-14 book/paper.tex
Moved book code into code folder

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{triangle.cxl}
\end{code}
\caption{Computing triangle numbers.}
\label{fig:triangle}
\end{figure}

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{Up.cxl}
\end{code}
\caption{Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{csbarebones.cxl}
\end{code}
\caption{A barebones critical section with two processes.}
\label{fig:csbarebones}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{cs.cxl}
\end{code}
\caption{CXL model of a critical section.}
\label{fig:cs}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpLock.cxl}
\end{code}
\caption{Na\"{\i}ve implementation of a shared lock.}
\label{fig:uplock}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFlags.cxl}
\end{code}
\caption{Na\"{\i}ve use of flags to solve mutual exclusion.}
\label{fig:upflags}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTurn.cxl}
\end{code}
\caption{Na\"{\i}ve use of turn variable to solve mutual exclusion.}
\label{fig:upturn}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../Peterson.cxl}
\end{code}
\caption{Peterson's Algorithm}
\label{fig:peterson}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonProof.cxl}
\end{code}
\caption{Peterson's Algorithm with Inductive Invariant}
\label{fig:petersonproof}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonMethod.cxl}
\end{code}
\caption{Peterson's Algorithm accessed through methods.}
\label{fig:petersonmethods}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTAS.cxl}
\end{code}
\caption{Mutual Exclusion using a ``spinlock'' based on test-and-set.}
\label{fig:tas}
\end{figure}

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTASinv.cxl}
\end{code}
\caption{Checking invariants.}
\label{fig:tasinv}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../TAS.cxl}
\end{code}
\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
\label{fig:tas2}
\end{figure}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
def tas(lk):
    atomic:
        result = ^lk;
        ^lk = True;
    ;
;
def Lock():
    result = False;
;
def lock(lk):
    while tas(lk):
        pass;
    ;
;
def unlock(lk):
    ^lk = False;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
\label{fig:spinlocks}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFixed.cxl}
\end{code}
\caption{Program of Figure~\ref{fig:inc} fixed with a lock.}
\label{fig:incfixed}
\end{figure}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
import list;

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbusy.cxl}
\end{code}
\caption{Busy-Waiting Reader/Writer Lock implementation.}
\label{fig:rwbusy}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWblock.cxl}
\end{code}
\caption{Checking for Busy Waiting.}
\label{fig:rwblock}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWlock.cxl}
\end{code}
\caption{Reader/Writer with Two Locks.}
\label{fig:rw2lock}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWmulti.cxl}
\end{code}
\caption{Checking that multiple readers can acquire the read lock.}
\label{fig:rwmulti}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../Diners.cxl}
\end{code}
\caption{Dining Philosophers.}
\label{fig:diners}
\end{figure}

@@

\begin{code}
\begin{verbatim}
    if left < right:
        call lock &(forks[left]);
        call lock &(forks[right]);
    else:
        call unlock &(forks[right]);
        call unlock &(forks[left]);
    ;
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
def Semaphore(cnt):
    result = cnt;
;
def P(sema):
    let blocked = True:
        while blocked:
            atomic:
                if (^sema) > 0:
                    ^sema = (^sema) - 1;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def V(sema):
    atomic:
        ^sema = (^sema) + 1;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../DinersSema.cxl}
\end{code}
\caption{Italian Philosophers.}
\label{fig:dinerssema}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsema.cxl}
\end{code}
\caption{Bounded Buffer implementation using semaphores.}
\label{fig:boundedbuffer}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsemadata.cxl}
\end{code}
\caption{Test whether the bounded buffer delivers the right data in the
right order.}
\label{fig:PCsemadata}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
\end{code}
\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
\label{fig:RWsplitsema}
\end{figure}

@@

CCSs are easy to implement using busy waiting:
\begin{code}
\begin{verbatim}
let blocked = True:
    while blocked:
        call lock();
        if application-condition-holds:
            blocked = False;
        else:
            call unlock():
@cr: application-code;
call unlock():
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWfair.cxl}
\end{code}
\caption{Reader/Writer Lock implementation addressing fairness.}
\label{fig:RWfair}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWcv.cxl}
\end{code}
\caption{Unfair Reader/Writer Lock implemtation using Mesa-style condition variables.}
\label{fig:RWcv}
\end{figure}

>>>>

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/triangle.cxl}
\end{code}
\caption{[triangle.cxl] Computing triangle numbers.}
\label{fig:triangle}
\end{figure}

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.cxl}
\end{code}
\caption{[Up.cxl] Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csbarebones.cxl}
\end{code}
\caption{[csbarebones.cxl] A barebones critical section with two processes.}
\label{fig:csbarebones}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/cs.cxl}
\end{code}
\caption{[cx.cxl] CXL model of a critical section.}
\label{fig:cs}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveLock.cxl}
\end{code}
\caption{[naiveLock.cxl] Na\"{\i}ve implementation of a shared lock.}
\label{fig:uplock}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveFlags.cxl}
\end{code}
\caption{[naiveFlags.cxl] Na\"{\i}ve use of flags to solve mutual exclusion.}
\label{fig:upflags}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveTurn.cxl}
\end{code}
\caption{[naiveTurn.cxl] Na\"{\i}ve use of turn variable to solve mutual exclusion.}
\label{fig:upturn}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Peterson.cxl}
\end{code}
\caption{[Peterson.cxl] Peterson's Algorithm}
\label{fig:peterson}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonInductive.cxl}
\end{code}
\caption{[PetersonInductive.cxl] Peterson's Algorithm with Inductive Invariant}
\label{fig:petersonproof}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonMethod.cxl}
\end{code}
\caption{[PetersonMethod.cxl] Peterson's Algorithm accessed through methods.}
\label{fig:petersonmethods}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlock.cxl}
\end{code}
\caption{[spinlock.cxl] Mutual Exclusion using a ``spinlock'' based on test-and-set.}
\label{fig:tas}
\end{figure}

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlockInv.cxl}
\end{code}
\caption{[spinlockInv.cxl] Checking invariants.}
\label{fig:tasinv}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csTAS.cxl}
\end{code}
\caption{[csTAS.cxl] Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
\label{fig:tas2}
\end{figure}

@@

\begin{figure}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def tas(lk):
    atomic:
        result = ^lk;
        ^lk = True;
    ;
;
def Lock():
    result = False;
;
def lock(lk):
    while tas(lk):
        pass;
    ;
;
def unlock(lk):
    ^lk = False;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
\label{fig:spinlocks}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpLock.cxl}
\end{code}
\caption{[UpLock.cxl] Program of Figure~\ref{fig:inc} fixed with a lock.}
\label{fig:incfixed}
\end{figure}

@@

\begin{figure}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
import list;

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusy.cxl}
\end{code}
\caption{[RWbusy.cxl] Busy-Waiting Reader/Writer Lock implementation.}
\label{fig:rwbusy}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusychk.cxl}
\end{code}
\caption{[RWbusychk.cxl] Checking for Busy Waiting.}
\label{fig:rwblock}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWlock.cxl}
\end{code}
\caption{[RWlock.cxl] Reader/Writer with Two Locks.}
\label{fig:rw2lock}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWmulti.cxl}
\end{code}
\caption{[RWmulti.cxl] Checking that multiple readers can acquire the read lock.}
\label{fig:rwmulti}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Diners.cxl}
\end{code}
\caption{[Diners.cxl] Dining Philosophers.}
\label{fig:diners}
\end{figure}

@@

\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
    if left < right:
        call lock &(forks[left]);
        call lock &(forks[right]);
    else:
        call unlock &(forks[right]);
        call unlock &(forks[left]);
    ;
\end{Verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def Semaphore(cnt):
    result = cnt;
;
def P(sema):
    let blocked = True:
        while blocked:
            atomic:
                if (^sema) > 0:
                    ^sema = (^sema) - 1;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def V(sema):
    atomic:
        ^sema = (^sema) + 1;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersSema.cxl}
\end{code}
\caption{[DinersSema.cxl] Italian Philosophers.}
\label{fig:dinerssema}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsema.cxl}
\end{code}
\caption{[BBsema.cxl] Bounded Buffer implementation using semaphores.}
\label{fig:boundedbuffer}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsemadata.cxl}
\end{code}
\caption{[BBsemadata.cxl] Test whether the bounded buffer delivers the right data in the
right order.}
\label{fig:PCsemadata}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWsbs.cxl}
\end{code}
\caption{[RWsbs.cxl] Reader/Writer Lock implementation using Split Binary Semaphores.}
\label{fig:RWsplitsema}
\end{figure}

@@

CCSs are easy to implement using busy waiting:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
let blocked = True:
    while blocked:
        call lock();
        if application-condition-holds:
            blocked = False;
        else:
            call unlock():
@cr: application-code;
call unlock():
\end{Verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWfair.cxl}
\end{code}
\caption{[RWfair.cxl] Reader/Writer Lock implementation addressing fairness.}
\label{fig:RWfair}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWcv.cxl}
\end{code}
\caption{[RWcv.cxl] Reader/Writer Lock implementation using Mesa-style condition variables.}
\label{fig:RWcv}
\end{figure}
